---
title: "Statistical Analysis"
author: "Arghya Kannadaguli (ak5357)"
date: "2025-05-07"
output: 
  html_document:
    df_print: paged
    toc: yes
    toc_float: yes
    code_folding: hide
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(tidyverse)
library(dplyr)
library(ggplot2)
library(gt)
library(psych)
```

Import Data
```{r}
dat = read_csv("data/AAMEHS Final Dataset.csv") |> 
  suppressMessages()

dat |> nrow()
# dat |> glimpse()
# dat |> names()
```

Test OLS
```{r}
# Test OLS
dat.sub = dat |>
  select(low_birth_weight, drinking_water, avg_birthing_age,
         unemployment_rate, health_insurance_coverage, poverty_rate_families,
         edu_hs, fms_never_married, foreign_born)

dat.sub|> 
  lm(low_birth_weight ~ ., data = _) |> 
  broom::tidy() |> 
  arrange(p.value)

ggcorr(dat.sub)
```

Summary Table
```{r}
sum.vars = c("low_birth_weight", "drinking_water", "avg_birthing_age", "unemployment_rate", "poverty_rate_families", "edu_hs", "fms_never_married", "foreign_born", "limited_english")

sum.names = c("Low Birth Weight (%)", "Drinking Water Contaminant Indicator", "Average Birthing Age", "Unemployment Rate (%)", "Family Poverty Rate (%)", "High School Diploma or Certificate (%)", "Never Married (% Females)", "Foreign Born (%)", "Limited English Proficiency (%)")

sum.table = dat |> 
  select(all_of(sum.vars))

summary_table = tibble(
    #Variable = sum.vars[sum.vars %in% names(sum.table)],
    Term = sum.names[sum.vars %in% names(sum.table)],
    Mean = unname(sapply(sum.table, mean)),
    SD = unname(sapply(sum.table, sd)),
    Min = unname(sapply(sum.table, min)),
    Max = unname(sapply(sum.table, max))
)

summary_table_gt = summary_table |>
  gt() |>
  fmt_number(columns = c(Mean, SD, Min, Max), decimals = 1) |>
  tab_style(
    style = cell_text(weight = "bold"),
    locations = cells_column_labels(everything())) |>
  tab_header(
    title = "Table 1. Summary Statistics",
    subtitle = "Descriptive Statistics for Selected Variables") |>
  cols_label(
    Term = "Term",
    Mean = "Mean",
    SD = "SD",
    Min = "Min",
    Max = "Max") |>
  tab_options(
    table.font.size = "small",
    heading.align = "center")
summary_table_gt
```


```{r eval = FALSE}
gtsave(summary_table_gt, "figures/summary_table.png")
```

Get subsets of dataset
```{r eval = FALSE}
# Selected Covariates
dat.sub1 = dat |>
  select(low_birth_weight, drinking_water, avg_birthing_age,
         unemployment_rate, poverty_rate_families,
         edu_hs, fms_never_married, foreign_born, limited_english)

# avg_birthing_age

cat(dat.fa |> names() |> paste0(collapse = "\n")) 
```

Test OLS mdoels
```{r eval = FALSE}
dat.sub1 |> 
  lm(low_birth_weight ~ ., data = _) |> 
  broom::tidy() |> 
  arrange(p.value) |> 
  knitr::kable()
```

Prep data for FA
```{r}
# All possible covariates
dat.fa = dat |> 
  select(-c(geo_id:longitude, drinking_water, low_birth_weight, total_birth_pop, total_birth_count, birthing_rate, avg_birthing_age, foreign_born, limited_english, edu_hs, edu_some_college, edu_associate, edu_bachelors, edu_bachelors_or_higher, edu_grad_prof, poverty_rate_all)) |> 
  lapply(scale) |> # scale and center all values
  as_tibble()
```

Identify best FA nfactors parameter value
Appears to be 5 for smaller subset
```{r eval = FALSE}
fct = tibble(n_fact = 2:15, BIC = NA_real_)

for (i in 2:15){
  print(i)
  set.seed(2025)
  fa = fa(dat.fa, nfactors = i, rotate = "varimax", scores = "regression", fm = "ml")
  fct$BIC[fct$n_fact == i] = fa$BIC
}
rm(fa)

fct$label = NA
fct$label[fct$BIC == min(fct$BIC)] = "Minimum BIC"

fct |> 
  ggplot(aes(x = n_fact, y = BIC)) +
  geom_line(lwd = 0.2) +
  geom_point() +
  geom_point(aes(x = n_fact, y = BIC), data = filter(fct, n_fact == 5), color = "red") +
  geom_text(aes(label = n_fact), vjust = -0.8, size = 2) + 
  labs(
    title = "BIC by Number of Factors",
    x = "Number of Factors in Factor Analysis",
    y = "Bayesian Information Criterion (BIC)",
    color = "",
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = "bold"),
    axis.title = element_text(size = 10),
    legend.position = "none") + 
  theme_classic() +
  theme(plot.title = element_text(hjust = 0.5))
```

Prep Data for FA
```{r}
dat.fa1 = dat.sub1 |>
  select(-low_birth_weight, drinking_water, avg_birthing_age)
```


Perform FA
```{r}
fa.5 = fa(dat.fa, nfactors = 5, rotate = "promax", scores = "regression", fm = "ml")
loadings = data.frame(fa.5$loadings[])

ggcorr(fa.5$scores)
```

Scree Plot
```{r}
eigenvalues = fa.5$values
ggplot(data.frame(Factor = 1:length(eigenvalues), Variance = eigenvalues), 
       aes(x = Factor, y = Variance)) +
  geom_line(color = "red", size = 1) +
  geom_point(size = 3) +
  xlim(0,5) +
  labs(title = "Scree Plot for Factor Analysis",
       x = "Factor Number",
       y = "Eigenvalue (Variance)") +
  theme_minimal()


variance_explained <- fa.5$Vaccounted
print(variance_explained)
```

Visualize Factors
```{r}
sum.loadings = loadings |> 
  select(ML1, ML3, ML2) |> 
  rownames_to_column() |> 
  as_tibble() |> 
  rename(constituent = rowname) |> 
  pivot_longer(
    -constituent,
    names_to = "factor",
    values_to = "measurement") |> 
  mutate(
    factor = str_replace(factor, "ML", "Factor "),
    label = case_match(
      constituent,
      "teen_birth" ~ "Teen Birth",
      "geriatric_birth" ~ "Geriatric Birth",
      "female_employment_rate" ~ "Female Employment Rate",
      "unemployment_rate" ~ "Unemployment Rate",
      "health_insurance_coverage" ~ "Health Insurance Coverage",
      "disability" ~ "Disability",
      "median_income" ~ "Median Income",
      "poverty_rate_families" ~ "Poverty Rate",
      "residence_over_1y" ~ "Residential Stability",
      "edu_less_than_hs" ~ "Less than High School",
      "edu_some_hs" ~ "Some High School",
      "edu_hs_or_higher" ~ "High School or Higher",
      "fms_never_married" ~ "FMS Never Married",
      "fms_married_separated" ~ "FMS Maried and Separated",
      "fms_separated" ~ "FMS Separated",
      "fms_widowed" ~ "FMS Widowed",
      "fms_divorced" ~ "FMS Divorced",
      "foreign_born_europe" ~ "FBO Europe",
      "foreign_born_asia" ~ "FBO Asia",
      "foreign_born_africa" ~ "FBO Africa",
      "foreign_born_oceania" ~ "FBO Oceania",
      "foreign_born_latin_america" ~ "FBO Latin America",
      "foreign_born_north_america" ~ "FBO North America",
      "foreign_born_not_citizen" ~ "FB Non-Citizen",
      "limited_english_spanish" ~ "LEP Spanish",
      "limited_english_indo_euro" ~ "LEP Indo-European",
      "limited_english_aapi" ~ "LEP Asian or Pacific Islander",
      "limited_english_other" ~ "LEP Other"),
    id = case_match(
      constituent,
      "teen_birth" ~ 1,
      "geriatric_birth" ~ 2,
      "female_employment_rate" ~ 3,
      "unemployment_rate" ~ 4,
      "health_insurance_coverage" ~ 5,
      "disability" ~ 6,
      "median_income" ~ 7,
      "poverty_rate_families" ~ 8,
      "residence_over_1y" ~ 9,
      "edu_less_than_hs" ~ 10,
      "edu_some_hs" ~ 11,
      "edu_hs_or_higher" ~ 12,
      "fms_never_married" ~ 13,
      "fms_married_separated" ~ 14,
      "fms_separated" ~ 15,
      "fms_widowed" ~ 16,
      "fms_divorced" ~ 17,
      "foreign_born_europe" ~ 18,
      "foreign_born_asia" ~ 19,
      "foreign_born_africa" ~ 20,
      "foreign_born_oceania" ~ 21,
      "foreign_born_latin_america" ~ 22,
      "foreign_born_north_america" ~ 23,
      "foreign_born_not_citizen" ~ 24,
      "limited_english_spanish" ~ 25,
      "limited_english_indo_euro" ~ 26,
      "limited_english_aapi" ~ 27,
      "limited_english_other" ~ 28),
    label = paste0(id, ". ", label),
    label = fct_reorder(.f = label, .x = id, .fun = min),
    constituent = fct_reorder(.f = constituent, .x = id, .fun = min)
    )

my.colors = c(
  unname(pals::stepped()),
  unname(pals::stepped())
)

fa.plot = sum.loadings |> 
  ggplot(aes(x = label, y = measurement, fill = label)) +
  geom_bar(stat = "identity") +
  facet_wrap(~ factor, nrow = 1) +
  #ylim(NA, 10.5) +
  #scale_color_manual(values = stepped(20)) +
  scale_fill_manual(values = my.colors) +
  theme_classic() +
  geom_text(aes(label = id,
                vjust = ifelse(measurement > 0, -0.5, 1)
                ), size.unit = "mm", size = 3) +
  labs(
    title = "Factor Analysis by Socioeconomic Variables",
    x = "",
    y = "Scaled and Centered Concentration",
    fill = "SES Variable",
    caption = "FB = Foreign Born, FBO = Foreign Born Origin, FMS = Female Marital Status, LEP = Limited English Proficiency"
  ) +
  theme(
    legend.position = "bottom",
    axis.text.x = element_blank(),#element_text(vjust = 1, hjust = 1),
    axis.ticks.x = element_blank(),
    plot.title = element_text(face = "bold", hjust = 0.5))
fa.plot

ggsave("figures/fa_plot.png", plot = fa.plot, width = 11, height = 6.5, units = "in")

sum.loadings$constituent |> unique()
```


```{r eval = FALSE}
library(RColorBrewer)
# display.brewer.all()
# sum.loadings$constituent |> unique()

my.colors1 = c(
  brewer.pal(n = 3, name = "Reds")[2:3], #teen, ger
  brewer.pal(n = 3, name = "Oranges")[2:3], #femp, unemp
  brewer.pal(n = 3, name = "Greens")[2:3], #insurance, disability
  brewer.pal(n = 4, name = "Blues")[2:4], #income, poverty, residence
  brewer.pal(n = 4, name = "RdPu")[2:4], #edu x3
  brewer.pal(n = 6, name = "PuBuGn")[2:6], #fms x5
  brewer.pal(n = 8, name = "Oranges")[2:8], #fb x7
  brewer.pal(n = 5, name = "Oranges")[2:5] #lep x4
)

my.colors2 = c(
  brewer.pal(n = 9, name = "PuBuGn"), #indiv x9
  brewer.pal(n = 3, name = "RdPu"), #edu x3
  brewer.pal(n = 5, name = "YlGnBu"), #fms x5
  brewer.pal(n = 7, name = "YlGnBu"), #fb x7
  brewer.pal(n = 4, name = "YlGnBu") #lep x4
)
#indiv x9 (teen, ger, femp, unemp, insurance, disability, income, poverty, residence)

my.colors =  c(
  brewer.pal(n = 11, name = "Spectral"),
  brewer.pal(n = 11, name = "Spectral"),
  brewer.pal(n = 11, name = "Spectral")
)
length(my.colors)
length(unique(sum.loadings$constituent))
```


teen_birth, geriatric_birth,
female_employment_rate, unemployment_rate, health_insurance_coverage,
disability,
median_income,
poverty_rate_families,
residence_over_1y,
edu_less_than_hs,
edu_some_hs,
edu_hs_or_higher,
fms_never_married,
fms_married_separated,
fms_separated,
fms_widowed,
fms_divorced,
foreign_born_europe,
foreign_born_asia,
foreign_born_africa,
foreign_born_oceania,
foreign_born_latin_america,
foreign_born_north_america,
foreign_born_not_citizen,
limited_english_spanish,
limited_english_indo_euro,
limited_english_aapi,
limited_english_other

FA Regression

```{r}
loadings

dat.fareg = dat |> 
  select(low_birth_weight, drinking_water, avg_birthing_age) |> 
  bind_cols(as_tibble(fa.5$scores[,1:3]))

lm(low_birth_weight ~ ., data = dat.fareg) |> 
  broom::tidy() |> 
  mutate(
    ci.ll = estimate - 1.96*std.error,
    ci.ul = estimate + 1.96*std.error) |> 
  select(term, estimate, p.value, contains("ci")) |> 
  arrange(p.value)

loadings |> 
  arrange(-abs(ML3)) |> 
  select(ML3)

```



Get Results
```{r}
summary(fa.sub1.5)
summary(fa.sub2.5)
summary(fa.sub2.10)

library(GGally)
ggcorr(dat.sub)
ggcorr(dat.sub2)

```



